Improving Haskell Theory Exploration

### 2015-12-04 TransferReport/contributions.tex
Fleshed out flattening algorithm

To achieve this we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}. Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$. $n_T$ may differ for each (sub-) tree for each sub-tree; trees where $n_T = 0$ are \emph{leaves}. Rose trees for the Peano arithmetic functions of figure \ref{fig:coreexample} are given in figure \ref{fig:rosetreeexample}.

@@

To flatten a rose tree we fetch the features from each ``level'', concatenating those from adjacent sub-trees:

\begin{equation*}
  level(l, (x, T_1, \dots, T_{n_T})) =
    \begin{cases}
      (x) & \text{if $l = 1$} \\
      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$ and $\vect{x}_i = level(l - 1, T_i)$}
    \end{cases}
\end{equation*}

Given a rose tree $T$, we pad each level $level(1, T)$, $level(2, T)$, $\dots$, $level(r, T)$ with zeros and truncate them to length $c$. The resulting vectors are concatenated to form the feature vector for $T$.

See see figure \ref{fig:rosetreeexample} for an example of this construction.

\begin{figure}
  \begin{small}
  \begin{align*}
    toTree(\hs{plus}) &= \\
      (\feature{\CLam},
       \begin{array}[t]{@{}l@{}}
         (\feature{\id{a}}), \\
         (\feature{\CLam},
          \begin{array}[t]{@{}l@{}}
            (\feature{\id{y}}), \\
            (\feature{\CCase},
             \begin{array}[t]{@{}l@{}}
               (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
               (\feature{\id{b}}), \\
               (\feature{\CAlt},
                \begin{array}[t]{@{}l@{}}
                  (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
                  (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}})))),
                \end{array} \\
               (\feature{\CAlt},
                \begin{array}[t]{@{}l@{}}
                  (\feature{\CDataAlt}, (\feature{\id{S}})), \\
                  (\feature{\CApp},
                   \begin{array}[t]{@{}l@{}}
                     (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{S}}))), \\
                     (\feature{\CApp},
                      \begin{array}[t]{@{}l@{}}
                        (\feature{\CApp},
                         \begin{array}[t]{@{}l@{}}
                           (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{plus}}))), \\
                           (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{x}})))),
                         \end{array} \\
                      (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{y}}))))),
                      \end{array}
                   \end{array} \\
                  (\feature{\id{x}})))))
                \end{array}
             \end{array}
          \end{array}
       \end{array} \\
\iffalse
    toTree(\hs{mult}) &= \\
      (Lam "a" (Lam "y" (Case (Var (Local "a"))
                              "b"
                              [Alt (DataAlt "Z") (Var (Global "Z")) [],
                               Alt (DataAlt "S") (App (App (Var (Global "plus"))
                                                           (Var (Local  "y")))
                                                      (App (App (Var (Global "mult"))
                                                                (Var (Local  "x")))
                                                           (Var (Local  "y"))))
                                                 ["x"]])))

    toTree(\hs{odd}) &= \\
      (\feature{\CLam},
       \begin{array}[t]{@{}l@{}}
         (\feature{\id{a}}), \\
         (\feature{\CCase},
          \begin{array}[t]{@{}l@{}}
            (\feature{\CVar}, (\feature{\CLocal}, (\feature{\id{a}}))), \\
            (\feature{\id{b}}), \\
            (\feature{\CAlt},
             \begin{array}[t]{@{}l@{}}
               (\feature{\CDataAlt}, (\feature{\id{Z}})), \\
               (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{False}})))),
             \end{array} \\
            (\feature{\CAlt},
             \begin{array}[t]{@{}l@{}}
               (\feature{\CDataAlt}, (\feature{\id{S}})), \\
               (\feature{\CApp},
                \begin{array}[t]{@{}l@{}}
                  (\feature{\CVar}, (\feature{\CGlobal}, (\feature{\id{even}}))), \\
                  (\feature{\CVar}, (\feature{\CLocal},  (\feature{\id{n}})))),
                \end{array} \\
               (\feature{\id{n}}))))
             \end{array}
          \end{array}
       \end{array}
\fi

    toTree(\hs{odd}) &= \\
      \begin{array}
        (\feature{\CLam}, &(\feature{\id{a}}),&                    &                      &                      &                           &                         \\
                          &(\feature{\CCase}, &(\feature{\CVar},   &(\feature{\CLocal},   &(\feature{\id{a}}))), &                           &                         \\
                          &                   &(\feature{\id{b}}), &                      &                      &                           &                         \\
                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{Z}})),  &                           &                         \\
                          &                   &                    &(\feature{\CVar},     &(\feature{\CGlobal},  &(\feature{\id{False}})))), &                         \\
                          &                   &(\feature{\CAlt},   &(\feature{\CDataAlt}, &(\feature{\id{S}})),  &                           &                         \\
                          &                   &                    &(\feature{\CApp},     &(\feature{\CVar},     &(\feature{\CGlobal},       &(\feature{\id{even}}))), \\
                          &                   &                    &                      &(\feature{\CVar},     &(\feature{\CLocal},        &(\feature{\id{n}})))),   \\
                          &                   &                    &(\feature{\id{n}})))) &                      &                           &

       \end{array}
  \end{align*}
  \end{small}
  \caption{}
  \label{fig:rosetreeexample}
\end{figure}

>>>>

To achieve this we first transform the nested tokens of an expression into a \emph{rose tree} of features, using the $toTree$ function shown in figure \ref{fig:totree}. Rose trees are defined recursively: $T$ is a rose tree if $T = (f, T_1, \dots, T_{n_T})$, where $f \in \mathbb{R}$ and $T_i$ are rose trees. $T_i$ are the \emph{sub-trees} of $T$ and $f$ is the \emph{feature at} $T$. $n_T$ may differ for each (sub-) tree; trees where $n_T = 0$ are \emph{leaves}. The results are illustrated in figure \ref{fig:rosetreeexample}.

@@

\begin{figure}
  \centering
  \begin{subfigure}{\textwidth}
    \begin{small}
      \Tree[ .$\feature{\CLam}$
                $\feature{\id{a}}$
                [ .$\feature{\CCase}$
                     [ .$\feature{\CVar}$
                          [ .$\feature{\CLocal}$
                               $\feature{\id{a}}$ ]]
                     $\feature{\id{b}}$
                     [ .$\feature{\CAlt}$
                          [ .$\feature{\CDataAlt}$
                               $\feature{\id{Z}}$ ]
                          [ .$\feature{\CVar}$
                               [ .$\feature{\CGlobal}$
                                    $\feature{\id{False}}$ ]]]
                     [ .$\feature{\CAlt}$
                          [ .$\feature{\CDataAlt}$
                               $\feature{\id{S}}$ ]
                          [ .$\feature{\CApp}$
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CGlobal}$
                                         $\feature{\id{even}}$ ]]
                               [ .$\feature{\CVar}$
                                    [ .$\feature{\CLocal}$
                                         $\feature{\id{n}}$ ]]]
                          $\feature{\id{n}}$ ]]]
    \end{small}
    \caption{Rose tree for the expression \hs{odd} from \ref{fig:coreexample}. Each (sub-) rose tree is rendered with its feature at the node and sub-trees beneath.}
    \label{fig:rosetreeexample}
  \end{subfigure}
  \vspace{1ex}
  \begin{subfigure}{\textwidth}
    \begin{equation*}
      \begin{bmatrix}
        \feature{\CLam}      & 0                   & 0                  & 0                   & 0               & 0                \\
        \feature{\id{a}}     & \feature{\CCase}    & 0                  & 0                   & 0               & 0                \\
        \feature{\CVar}      & \feature{\id{b}}    & \feature{\CAlt}    & \feature{\CAlt}     & 0               & 0                \\
        \feature{\CLocal}    & \feature{\CDataAlt} & \feature{\CVar}    & \feature{\CDataAlt} & \feature{\CApp} & \feature{\id{n}} \\
        \feature{\id{a}}     & \feature{\id{Z}}    & \feature{\CGlobal} & \feature{\id{S}}    & \feature{\CVar} & \feature{\CVar}  \\
        \feature{\id{False}} & \feature{\CGlobal}  & \feature{\CLocal}  & 0                   & 0               & 0                \\
        \feature{\id{even}}  & \feature{\id{n}}    & 0                  & 0                   & 0               & 0
      \end{bmatrix}
    \end{equation*}
    \caption{Matrix generated from figure \ref{fig:rosetreeexample}, padded to 6 columns. Each level of sub-trees corresponds to a row in the matrix.}
    \label{fig:matrixexample}
  \end{subfigure}
  \vspace{1ex}
  \begin{subfigure}{\textwidth}
    \begin{equation*}
      (\feature{\CLam}, 0, 0, 0, 0, 0, \feature{\id{a}}, \feature{\CCase}, 0, 0, 0, 0, \feature{\CVar}, \feature{\id{b}}, \feature{\CAlt}, \feature{\CAlt}, 0, 0, \dots
     \end{equation*}
     \caption{(Prefix of) the feature vector for \hs{odd}, constructed by concatenating the rows of \ref{fig:matrixexample}. Ignoring padding, the features are in breadth-first order.}
     \label{fig:vectorexample}
  \end{subfigure}
  \caption{Feature extraction applied to the expression \hs{odd} from figure \ref{fig:coreexample}.}
\end{figure}

We convert these rose trees into matrices, as shown in figure \ref{fig:matrixexample}, by gathering the features of adjacent (sub-) trees at each level of nesting using the following function:

\begin{equation*}
  level(l, (f, T_1, \dots, T_{n_T})) =
    \begin{cases}
      (f) & \text{if $l = 1$} \\
      \vect{x}_1 \vect{x}_2 \dots \vect{x}_{T_n} & \text{if $l > 1$, where $\vect{x}_i = level(l - 1, T_i)$}
    \end{cases}
\end{equation*}

Given a rose tree $t$ we can define its matrix $\mathbf{M}$ by $\mathbf{M}_i = pad(level(i, t))$, where $pad$ appends zeros to ensure each row has the same length. To improve alignment across different expressions, we actually pad each matrix to the same dimensions $r \times c$, which may require appending rows of zeros or truncating rows/columns as necessary.

Feature vectors are then simply the concatenation of matrix rows $(\mathbf{M}_{1,1} \mathbf{M}_{1,2} \dots \mathbf{M}_{2,1} \mathbf{M}_{2,2} \dots)$, as shown in figure \ref{fig:vectorexample}.
